\scshape Advanced game theory

### 2025-12-15 ms.tex
Add files via upload

\date{\small This version: 7 December 2025}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Measure and integral}
\label{math:meas}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The theory of measure and (Lebesgue) integration are the foundation of modern real analysis, which is in turn the backbone of economic theory.
You do not need to know it to take this course.
But it helps to know some of the basic \emph{language} of measure theory; that's what I'll cover here.

(If you'd like to pursue research in economic theory, then I would advise you to learn basic measure theory.
I taught myself from \textcite{Rosenthal2006}; this book is very accessible, except that chapter~2 is harder than it needs to be, so don't get stuck there! I now prefer the first few chapters of \textcite{Folland1999}, a very beautiful book for first-year graduate students in maths. There are lots of other standard books. Efe Ok has a `measure and probability' manuscript on his website that is specifically aimed at economists.)

Let $X$ be a non-empty set.
A \emph{$\sigma$-algebra} on $X$ is a collection of subsets of $X$ satisfying certain properties.%
	\footnote{The properties are closedness under complement and closedness under countable union.}
If $\mathcal{X}$ is a $\sigma$-algebra, we call $(X,\mathcal{X})$ a \emph{measurable space,} and call the elements of $\mathcal{X}$ the \emph{measurable subsets} of $X$.
Often the $\sigma$-algebra $\mathcal{X}$ is left partly or entirely implicit, by the way.

A \emph{measure} on a measurable space $(X,\mathcal{X})$ is a map $\mu : \mathcal{X} \to [0,\infty]$ that is countably additive:
for any countable collection $A_1,A_2,\dots$ of pairwise disjoint measurable subsets of $X$,
we have $\mu\left( \Union_{n \in \N} A_n \right) = \sum_{n \in \N} \mu(A_n)$.
The triple $(X,\mathcal{X},\mu)$ is called a \emph{measure space.}

A measurable set $A \subseteq X$ is called \emph{$\mu$-null} iff $\mu(A)=0$. If a property holds at every $x \in X$, except possibly for $x$ belonging to a null set $A$, then that property is said to hold \emph{($\mu$-)almost everywhere,} or `($\mu$-)a.e.'.

\begin{example}
	%
	\label{example:lebesgue_measure}
	%
	The most commonly-used measure on $X=\R$ is the \emph{Lebesgue measure,}
	which is the unique measure $\lambda$ with the property that
	$\lambda([a,b]) = b-a$ for all $a<b$.
	That is, it captures the common-sense notion of \emph{length.}

	Analogously, there's Lebesgue measure on $\R^2$, which captures \emph{area,} and Lebesgue measure on $\R^3$, which captures \emph{volume,} and so on.

	The Lebesgue measure is conventionally defined on the \emph{Lebesgue $\sigma$-algebra} (whose elements are called \emph{Lebesgue sets}).
	It is often easier to work with the coarser \emph{Borel $\sigma$-algebra} (or \emph{Borel sets}); this is the smallest $\sigma$-algebra containing every open subset of $\R$.
	%
\end{example}

If $\mu$ has the further property that $\mu(X)=1$, then it is a \emph{probability measure,}
and $(X,\mathcal{X},\mu)$ is a \emph{probability space.}
In the probability context, `almost everywhere' is usually replaced with `almost sure(ly)' or `a.s.'.

Now consider a function $f : X \to Y$, where both $X$ and $Y$ are measurable spaces. For any measurable set $B \subseteq Y$, a measure on $Y$ can tell us how `large' the set $B$ is.
How large, then, is the set $A = \{ x \in X : f(x) \in B \}$ of points $x$ in $X$ that lead to a value $f(x)$ that lives in $B$?
That question only has an answer if $A$ is a measurable set!
We call a function \emph{measurable} if this question has an answer for every set $B$.
Symbolically, $f$ is measurable iff for every measurable subset $B$ of $Y$, $\{ x \in X : f(x) \in B \}$ is a measurable subset of $X$.

In modern analysis, the standard integral is the Lebesgue integral.
This is the integral used almost exclusively in economic theory, including these notes.
The integral of a measurable function $f : X \to \R$ on a measure space $(X,\mathcal{X},\mu)$ is written $\int_X f \dd \mu$.
The integral over a measurable subset $A \subseteq X$ is defined
%
\begin{equation*}
	\int_A f \dd \mu \coloneqq \int_X f \1_A \dd \mu ,
\end{equation*}
%
where $\1_A(x) \coloneqq 1$ for $x \in A$ and $\coloneqq 0$ for $x \notin A$.

On probability spaces, measurable functions are conventionally called \emph{random variables,}
and integrals are called \emph{expectations.}
One writes
%
\begin{equation*}
	\E( f ) \coloneqq \int_X f \dd \mu .
\end{equation*}

If you have taken introductory calculus, then you may be more familiar with the \emph{Riemann integral.}%
	\footnote{The Riemann integral of a function $\R \to \R$ is defined as the limit of the areas of increasingly fine rectangles approximating the area under the graph of the function.}
The Riemann integral has the problem that the integral of some functions simply fail to exist.
The Lebesgue integral extends the Riemann integral:
it allows more functions to be integrated,
while still giving the same result as the Riemann integral whenever the latter exists.

The Lebesgue integral is defined only for functions $f : X \to \R$ that are measurable.
That is a necessary condition, but it is not sufficient:
the existence of the integral requires a further condition.
(If the further condition fails, the definition of the Lebesgue integral yields the expression $\infty - \infty$, which has no meaning; therefore we agree to say that the integral does not exist in such cases.)
The integral of a function may be infinite (equal to $\infty$ or $-\infty$).
A function is called \emph{integrable} iff both (a) its integral exists, \emph{and} (b) its integral is finite.

Whereas the Riemann integral is defined only for functions $f : \R \to \R$,
the Lebesgue integral makes sense for functions $f : X \to \R$
defined on any space $X$ you like, provided it has measurable structure (i.e. is equipped with a $\sigma$-algebra).
This is very useful.

>>>>

\date{\small This version: 15 December 2025}

@@
